---
title: "Elliptical graph layouts"
output: rmarkdown::html_vignette
bibliography: "`r system.file('references.bib', package='graphicalMCP')`"
vignette: >
  %\VignetteIndexEntry{Elliptical graph layouts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.dim = c(7, 4.2)
)
```

```{r setup, warning = FALSE, message = FALSE}
library(graphicalMCP)
```

# Overview

The default `plot()` method in `graphicalMCP` uses `igraph` to draw hypotheses
as vertices and transition weights as directed edges. The `layout = "ellipse"`
option keeps the same `igraph` plotting path, but arranges hypotheses around an
ellipse. This gives a communication-oriented display similar in spirit to the
elliptical arrangement in `gMCPLite::hGraph()`, while staying compatible with
the current `plot.initial_graph()` and `plot.updated_graph()` interface.

Elliptical layouts are useful when:

* the order of hypotheses matters and should be stable across plots;
* the graph has cycles or several bidirectional edges, where force-directed
  layouts can be unstable;
* a wider or taller plotting area would make labels and edge weights easier to
  read;
* hypotheses naturally fall into groups that can be shown with color.

The default elliptical layout is often a quick way to get a usable graph without
substantial hand tuning. It uses a stable ordering, separates hypotheses around
the plot boundary, and avoids the variability of force-directed layouts. The
main tradeoff is that an ellipse is not topology-aware. It does not try to
minimize edge crossings or edge-label overlap. For very dense graphs, large
graphs, or graphs with many long-range transitions, a custom layout matrix can
still be clearer. The ellipse layout is best viewed as a reproducible
presentation layout that often works well immediately, with options available
when a particular graph needs adjustment.

# A Simple Five-Hypothesis Graph

The example below has five hypotheses. The first two hypotheses start with most
of the testing weight, and rejected hypotheses can pass weight to later
hypotheses through the transition matrix.

```{r create-five-hypothesis-graph}
hypotheses <- c(
  H1 = 0.40,
  H2 = 0.25,
  H3 = 0.20,
  H4 = 0.10,
  H5 = 0.05
)

transitions <- rbind(
  H1 = c(0, 0.5, 0.5, 0, 0),
  H2 = c(0, 0, 0.4, 0.6, 0),
  H3 = c(0.3, 0, 0, 0, 0.7),
  H4 = c(0, 0, 0, 0, 1),
  H5 = c(0.5, 0.5, 0, 0, 0)
)

five_graph <- graph_create(hypotheses, transitions)
```

```{r simple-ellipse-plot}
plot(
  five_graph,
  layout = "ellipse",
  xradius = 2.7,
  yradius = 1.35,
  vertex.size = 72,
  vertex.size2 = 44,
  vertex.label.cex = 0.9,
  edge.arrow.size = 0.7
)
```

`xradius` and `yradius` control the shape of the layout ellipse. The vertices
also use an ellipse shape by default when `layout = "ellipse"`. Use
`vertex.size` for the horizontal radius of the vertex and `vertex.size2` for
the vertical radius. Larger hypothesis ellipses are helpful when labels include
both hypothesis names and weights.

# Grouped Hypotheses With A Legend

Because `plot()` still uses `igraph`, the usual `igraph::plot.igraph()`
arguments can be used. Here, color distinguishes two groups of hypotheses. The
legend is added with base R graphics after the graph is drawn. In this example,
`xlim` reserves space on the right edge for the legend.

```{r grouped-ellipse-plot}
hypothesis_groups <- c(
  H1 = "Primary",
  H2 = "Primary",
  H3 = "Secondary",
  H4 = "Secondary",
  H5 = "Secondary"
)

group_colors <- c(
  Primary = "#4477AA",
  Secondary = "#CC6677"
)

plot(
  five_graph,
  layout = "ellipse",
  xradius = 2.65,
  yradius = 1.3,
  xlim = c(-3.2, 5),
  ylim = c(-1.9, 1.9),
  vertex.color = group_colors[hypothesis_groups],
  vertex.size = 72,
  vertex.size2 = 44,
  vertex.label.cex = 0.9,
  edge.arrow.size = 0.7
)

legend(
  x = 3.4,
  y = 1.4,
  legend = names(group_colors),
  fill = group_colors,
  border = "black",
  title = "Group",
  bty = "n",
  cex = 0.85
)
```

# Custom Labels Inside Nodes

By default, nodes are labeled with the hypothesis name and hypothesis weight.
Pass `vertex.label` to use custom text instead. Line breaks (`\n`) are useful
when labels need to include a short endpoint name or other trial-specific
description.

```{r custom-node-labels}
custom_labels <- c(
  H1 = "ORR\n0.40",
  H2 = "PFS\n0.25",
  H3 = "OS\n0.20",
  H4 = "QoL\n0.10",
  H5 = "Safety\n0.05"
)

plot(
  five_graph,
  layout = "ellipse",
  xradius = 2.8,
  yradius = 1.35,
  vertex.label = custom_labels,
  vertex.size = 84,
  vertex.size2 = 48,
  vertex.label.cex = 0.85,
  edge.arrow.size = 0.7,
  edge_curves = c(default = 0.05, pairs = 0.3)
)
```

For longer labels, increase `vertex.size` and `vertex.size2`, decrease
`vertex.label.cex`, or both.

# Adjusting Arrow Curvature

By default, bidirectional edge pairs get a small curvature so the two arrows do
not sit directly on top of each other. With `edge_curves`, you can also set a
small default curvature for all non-paired edges, make paired arrows more
curved, or tune individual edges by name.

```{r curved-ellipse-plot}
plot(
  five_graph,
  layout = "ellipse",
  xradius = 2.8,
  yradius = 1.2,
  vertex.size = 72,
  vertex.size2 = 42,
  vertex.label.cex = 0.9,
  edge.arrow.size = 0.7,
  edge_curves = c(
    default = 0.08,
    pairs = 0.35,
    "H3|H1" = -0.18
  )
)
```

The named edge entries follow the convention `"tail|head"`. In this example,
`"H3|H1"` adjusts only the arrow from `H3` to `H1`.

# Changing The Starting Position

The first hypothesis is placed near the top by default, following the convention
used by `gMCPLite::hGraph()`. Use `radian_start` to rotate the layout, and
`clockwise = FALSE` if a counter-clockwise ordering is more natural for a
specific graph. This is analogous to the `radianStart` argument in
`gMCPLite::hGraph()`.

```{r rotated-ellipse-plot}
plot(
  five_graph,
  layout = "ellipse",
  xradius = 2.7,
  yradius = 1.35,
  radian_start = 0,
  clockwise = FALSE,
  vertex.size = 72,
  vertex.size2 = 44,
  vertex.label.cex = 0.9,
  edge.arrow.size = 0.7,
  edge_curves = c(default = 0.05, pairs = 0.3)
)
```

# Updated Graphs

`plot.updated_graph()` delegates to `plot.initial_graph()`, so the ellipse
layout can also be used after hypotheses are deleted from the graph. Deleted
hypotheses are shown in gray by default.

```{r updated-ellipse-plot}
updated_five_graph <- graph_update(
  five_graph,
  delete = c(TRUE, FALSE, FALSE, TRUE, FALSE)
)

plot(
  updated_five_graph,
  layout = "ellipse",
  xradius = 2.7,
  yradius = 1.35,
  vertex.size = 72,
  vertex.size2 = 44,
  vertex.label.cex = 0.9,
  edge.arrow.size = 0.7,
  edge_curves = c(default = 0.08, pairs = 0.35)
)
```

# A Top-First Ellipse Layout

The earlier examples use the default starting angle, which follows the
`gMCPLite::hGraph()` convention. If you want the first hypothesis at the top of
the ellipse, set `radian_start = pi / 2`. This gives a useful top-first layout
without supplying a custom coordinate matrix.

```{r top-first-ellipse-layout}
plot(
  five_graph,
  layout = "ellipse",
  xradius = 2.7,
  yradius = 1.35,
  radian_start = pi / 2,
  xlim = c(-3.2, 5),
  ylim = c(-1.9, 1.9),
  vertex.color = group_colors[hypothesis_groups],
  vertex.size = 72,
  vertex.size2 = 44,
  vertex.label.cex = 0.9,
  edge.arrow.size = 0.7,
  edge_curves = c(default = 0.06, pairs = 0.3)
)

legend(
  x = 3.4,
  y = 1.45,
  legend = names(group_colors),
  fill = group_colors,
  border = "black",
  title = "Group",
  bty = "n",
  cex = 0.85
)
```

# When To Use A Custom Layout Instead

The ellipse layout is meant to be a strong default for presentation. If a graph
has many transition weights, many edge crossings, or a trial-specific ordering
that cannot be expressed by rotating or reversing the ellipse, supply a
coordinate matrix directly to `layout`. This keeps full control over placement
while still using the same labeling, coloring, and edge-curvature tools.
