---
title: "Group Sequential Design with Graphical Approaches Using gsDesign"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Group Sequential Design with Graphical Approaches Using gsDesign}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 192,
  fig.retina = 2
)
```

## Introduction

Group sequential designs allow for interim analyses during a clinical trial,
enabling early stopping for efficacy while controlling the familywise Type I
error rate (FWER). When combined with graphical multiple comparison procedures,
these designs can test multiple hypotheses across multiple analyses in a
structured and transparent way.

The `graph_test_shortcut_gsDesign()` function performs the same group
sequential graphical procedure as `graph_test_shortcut_gsd()`, but obtains
every group sequential quantity — cumulative spending, boundaries, and repeated
p-values — from the [gsDesign package](https://CRAN.R-project.org/package=gsDesign)
(Anderson, 2026). Spending functions and spending time are specified in
gsDesign's own terms (`sfu`, `sfupar`, and `usTime`), so that a group
sequential design worked out with gsDesign can be carried over to the
graphical procedure unchanged. The graphical part of the procedure — the
shortcut testing, graph updates, and the `look_back` option — is identical to
`graph_test_shortcut_gsd()`, and the results have the same structure.

This vignette reproduces the two case studies of the
[Group Sequential Design with Graphical Approaches](group-sequential-testing.html)
vignette with `graph_test_shortcut_gsDesign()`: a diabetes trial from Section
4 of Maurer and Bretz (2013), and an oncology trial adapted from the
[gMCPLite package](https://cran.r-project.org/package=gMCPLite) where
different endpoints have different numbers of analyses. Along the way it shows
how gsDesign is used inside the function. gsDesign is a suggested package of
graphicalMCP and must be installed to use `graph_test_shortcut_gsDesign()`.

```{r gsdesign-guard, echo = FALSE, results = "asis"}
if (!requireNamespace("gsDesign", quietly = TRUE)) {
  cat(
    "**The gsDesign package is not installed, so the code in this vignette",
    "was not run.**\n"
  )
  knitr::knit_exit()
}
```

```{r setup, message = FALSE}
library(graphicalMCP)
library(gsDesign)
```

## Alpha Spending Functions

Alpha spending functions determine how the total significance level is allocated
across interim and final analyses. gsDesign provides a large collection of
spending functions. Each takes three arguments — `alpha` (total significance
level), `t` (spending time, usually the information fraction), and `param` (a
parameter, where applicable) — and returns a `spendfn` object whose `$spend`
element contains the cumulative alpha spent at each value of `t`. The most
common choices are:

- `sfLDOF()` — Lan-DeMets O'Brien-Fleming approximation (very conservative at
  early analyses)
- `sfLDPocock()` — Lan-DeMets Pocock approximation (more uniform spending)
- `sfHSD()` — Hwang-Shih-DeCani family (flexible, controlled by `param`)
- `sfPower()` — Kim-DeMets power family; `param = 1` gives linear (uniform)
  spending

```{r spending-comparison}
# Compare spending at information fractions (1/3, 2/3, 1) for alpha = 0.025
t_values <- c(1/3, 2/3, 1)
spending_comparison <- data.frame(
  `Info Fraction` = t_values,
  `O'Brien-Fleming` = sfLDOF(0.025, t_values)$spend,
  Pocock = sfLDPocock(0.025, t_values)$spend,
  `HSD (gamma=-4)` = sfHSD(0.025, t_values, param = -4)$spend,
  Linear = sfPower(0.025, t_values, param = 1)$spend,
  check.names = FALSE
)
knitr::kable(spending_comparison, digits = 6,
             caption = "Cumulative alpha spent at each analysis (alpha = 0.025)")
```

The O'Brien-Fleming spending function is very conservative at early analyses
(spending only `r sprintf("%.5f", sfLDOF(0.025, 1/3)$spend)` at the first
interim), making it a popular choice when early stopping should require very
strong evidence.

In `graph_test_shortcut_gsDesign()`, the spending function is passed as `sfu`
and its parameter as `sfupar`, exactly as in `gsDesign::gsDesign()`. The
built-in spending functions of graphicalMCP correspond to gsDesign's as
follows:

| graphicalMCP (`spending_fn`) | gsDesign (`sfu`, `sfupar`)       |
|------------------------------|----------------------------------|
| `spending_of()`              | `sfLDOF`                         |
| `spending_pocock()`          | `sfLDPocock`                     |
| `spending_hsd(gamma)`        | `sfHSD`, `sfupar = gamma`        |
| `spending_linear()`          | `sfPower`, `sfupar = 1`          |
| `spending_wt(delta)`         | `"WT"`, `sfupar = delta`         |
| `spending_with_time()`       | `usTime`                         |

## From Spending to Boundaries

An important distinction: the cumulative spending at an analysis is **not** the
nominal significance level (boundary) at that analysis. The boundary must be
computed from the spending using the joint distribution of test statistics across
analyses. Specifically, the Z-scale boundary $b_k$ satisfies:

$$P(Z_1 < b_1, \ldots, Z_k < b_k) = 1 - f(\alpha, t_k)$$

where $f(\alpha, t_k)$ is the cumulative spending and
$(Z_1, \ldots, Z_k)$ follow the canonical joint distribution with
$\text{Cor}(Z_i, Z_j) = \sqrt{t_i / t_j}$ for $i \le j$. We consider
one-sided tests with the upper alternative (i.e., larger effects are better).
At analysis $k$, the null hypothesis is rejected if $Z_k \ge b_k$, or
equivalently, if the observed p-value $p_k \le \Phi(-b_k)$ where $\Phi$ is
the standard normal CDF.

In gsDesign, the boundaries are obtained by constructing a design with
`gsDesign()`. The call below is the one that `graph_test_shortcut_gsDesign()`
makes for each hypothesis: `test.type = 1` requests a one-sided design with an
efficacy boundary only, and `n.I` together with `maxn.IPlan = 1` gives the
information at each analysis directly as a fraction of the planned maximum, so
that the correlation between analyses is the one implied by the information
fractions.

```{r boundaries-example}
# Compute boundaries for OBF spending at alpha = 0.025 with 3 equally spaced analyses
design_obf <- gsDesign(
  k = 3, test.type = 1, alpha = 0.025,
  sfu = sfLDOF, n.I = t_values, maxn.IPlan = 1
)

boundary_table <- data.frame(
  Analysis = 1:3,
  `Info Fraction` = t_values,
  `Cumulative Spending` = sfLDOF(0.025, t_values)$spend,
  `Z Boundary` = design_obf$upper$bound,
  `Nominal p Boundary` = pnorm(design_obf$upper$bound, lower.tail = FALSE),
  check.names = FALSE
)
knitr::kable(boundary_table, digits = 6,
             caption = "OBF group sequential boundaries (alpha = 0.025)")
```

## Repeated and Sequential P-values

Two types of p-values are central to the group sequential graphical procedure.
Let $\hat{p}_k$ denote the repeated p-value and $\tilde{p}_k$ denote the
sequential p-value at analysis $k$.

- **Repeated p-value** $\hat{p}_k$: the minimum significance level at which
  the group sequential boundary *at analysis $k$ specifically* would be crossed.
  It only considers the boundary at the current analysis.

- **Sequential p-value** $\tilde{p}_k$: the minimum significance level at
  which any group sequential boundary *at analyses $1, \ldots, k$* would be
  crossed. It equals the cumulative minimum of repeated p-values:
  $\tilde{p}_k = \min_{l=1}^{k} \hat{p}_l$.

gsDesign computes sequential p-values with `sequentialPValue()`, which takes a
design, the information (`n.I`) and Z-statistics (`Z`) at the analyses
conducted so far, and the spending time (`usTime`), and searches for the
smallest level at which any boundary is crossed. It does not provide repeated
p-values directly. `graph_test_shortcut_gsDesign()` obtains them from
`sequentialPValue()` with a simple device: the Z-statistics at analyses
$1, \ldots, k-1$ are replaced by $-20$. A Z-statistic of $-20$ can never cross
an efficacy boundary (gsDesign itself uses $-20$ internally as the value of
"no lower bound"), so the search then only involves the boundary at analysis
$k$ — which is precisely the definition of the repeated p-value. Because the
same design and the same integration are used, this is exact rather than an
approximation.

The following code shows the mechanism for a single hypothesis with the design
from the previous section. This is what `graph_test_shortcut_gsDesign()` does
internally for each hypothesis at each analysis; you never need to call
`sequentialPValue()` yourself.

```{r rep-seq-p-example}
# A hypothesis tested at three analyses with OBF spending
p_h1 <- c(0.05, 0.02, 0.01)
info <- c(1/3, 2/3, 1)
z_h1 <- qnorm(1 - p_h1)

# Sequential p-value at analysis 3 (considers boundaries at all three analyses)
seq_p_3 <- sequentialPValue(design_obf, n.I = info, Z = z_h1, usTime = info)
cat("Sequential p-value at analysis 3:", seq_p_3, "\n")

# Repeated p-value at analysis 3 (considers only the boundary at analysis 3):
# the earlier Z-statistics are replaced by -20
rep_p_3 <- sequentialPValue(
  design_obf, n.I = info, Z = c(-20, -20, z_h1[3]), usTime = info
)
cat("Repeated p-value at analysis 3:", rep_p_3, "\n")

# Repeated p-values at analyses 1 and 2 in the same way
rep_p_1 <- sequentialPValue(
  design_obf, n.I = info[1], Z = z_h1[1], usTime = info[1]
)
rep_p_2 <- sequentialPValue(
  design_obf, n.I = info[1:2], Z = c(-20, z_h1[2]), usTime = info[1:2]
)
cat("Repeated p-values:", rep_p_1, rep_p_2, rep_p_3, "\n")
cat("min(rep_p_1, rep_p_2, rep_p_3) =", min(rep_p_1, rep_p_2, rep_p_3), "\n")
```

The sequential p-value is the minimum of the repeated p-values across analyses
(up to the tolerance of gsDesign's root-finding). It is always less than or
equal to the repeated p-value, since it considers all prior analyses. A
hypothesis that nearly crossed its boundary at an earlier analysis will have a
much smaller sequential p-value than its repeated p-value at the current
analysis.

Note that the spending time `usTime` is passed explicitly in each call. When
it is omitted, `sequentialPValue()` rescales the spending time by the last
information supplied, which coincides with the information fraction here but
not when a trial over-runs its planned information (see the section on
monitoring below). `graph_test_shortcut_gsDesign()` therefore always passes
the spending time explicitly: `usTime` if supplied, and otherwise the
information fractions capped at 1.

The `graph_test_shortcut_gsDesign()` function supports two modes controlled by
the `look_back` parameter. When `look_back = FALSE` (the default), rejection
decisions at each analysis are based on repeated p-values only — i.e., only
the boundary at the current analysis is considered. When `look_back = TRUE`,
rejection decisions are based on sequential p-values, which "look back" at
all prior analyses by taking the cumulative minimum of repeated p-values.
This means that strong evidence from an earlier analysis is carried forward
and can contribute to a rejection at a later analysis. Both modes are
illustrated in the case studies below.

## Case Study using repeated p-values and look_back = FALSE

We replicate the numerical example from Section 4 of Maurer and Bretz (2013).
This is a diabetes trial with two preplanned interim analyses comparing two
doses (low and high) against placebo for two hierarchically ordered endpoints
(HbA1c and body weight).

### Hypotheses and Graph

The trial has four hypotheses:

- $H_1$: Low dose vs. placebo for HbA1c (primary)
- $H_2$: High dose vs. placebo for HbA1c (primary)
- $H_3$: Low dose vs. placebo for body weight (secondary)
- $H_4$: High dose vs. placebo for body weight (secondary)

The testing strategy follows the successiveness principle: secondary hypotheses
cannot be tested until their parent primary hypothesis is rejected. Both
primary hypotheses start with equal weight (0.5 each), and upon rejection,
the weight is split equally between the other primary hypothesis and the
corresponding secondary hypothesis.

```{r graph-setup}
hypotheses <- c(0.5, 0.5, 0, 0)
transitions <- rbind(
  c(0, 0.5, 0.5, 0), # H1 -> H2 (1/2), H3 (1/2)
  c(0.5, 0, 0, 0.5), # H2 -> H1 (1/2), H4 (1/2)
  c(0, 1, 0, 0),     # H3 -> H2 (full propagation)
  c(1, 0, 0, 0)      # H4 -> H1 (full propagation)
)
g <- graph_create(hypotheses, transitions)
```

```{r graph-plot, eval = requireNamespace("igraph", quietly = TRUE)}
plot(g, vertex.size = 40)
```

### Observed P-values

The trial has three planned analyses at equally spaced information fractions
$(1/3, 2/3, 1)$. The Lan-DeMets O'Brien-Fleming spending function `sfLDOF` is
used for all hypotheses. We assume the trial stops after the second interim
analysis (based on the results obtained). Since only two analyses are
conducted, we pass the p-values and information fractions for those two
analyses only.

```{r p-values}
# P-values from Table 1 of Maurer and Bretz (2013)
# Only analyses 1 and 2 are conducted (trial stops after analysis 2)
p <- rbind(
  H1 = c(0.0062, 0.0002),
  H2 = c(0.017,  0.0035),
  H3 = c(0.009,  0.002),
  H4 = c(0.13,   0.06)
)

p_display <- as.data.frame(p)
colnames(p_display) <- paste("Analysis", 1:2)
knitr::kable(
  p_display,
  caption = "Observed nominal p-values"
)
```

### Running the Procedure (look_back = FALSE)

The default mode is `look_back = FALSE`, which means the procedure does **not**
look back at test statistics from prior analyses. At each analysis $k$,
rejection decisions are based solely on the repeated p-value $\hat{p}_k$
computed from the test statistic at analysis $k$, without utilizing test
statistics from previous analyses. Note that the test statistic at analysis
$k$ is computed from all data accumulated up to that point, but the
rejection decision at analysis $k$ does not incorporate the test statistics
(or repeated p-values) from analyses $1, \ldots, k-1$.

There are two equivalent ways to understand the rejection decisions:

1. **Repeated p-values** (default): The repeated p-value $\hat{p}_k$ is the
   minimum significance level at which the group sequential boundary at
   analysis $k$ would be crossed. These are passed to the graphical shortcut
   procedure (`graph_test_shortcut()`) for multiplicity adjustment.
2. **Nominal boundaries** (`test_values = TRUE`): At each analysis, the
   observed p-value for each hypothesis is compared against its nominal
   boundary, which depends on the hypothesis's current weight in the graph
   and the spending function.

Both approaches lead to exactly the same rejection decisions — they are two
views of the same procedure.

#### Repeated p-values

`graph_test_shortcut_gsDesign()` computes the repeated p-value of each
hypothesis at each analysis from gsDesign as shown in the previous section,
and returns them in `$outputs$repeated_p`. The spending function is passed as
`sfu = sfLDOF`, which is also the default:

```{r run-gsd-no-lb}
info_frac <- c(1/3, 2/3)
result <- graph_test_shortcut_gsDesign(
  graph = g,
  p = p,
  alpha = 0.025,
  info_frac = info_frac,
  sfu = sfLDOF,
  look_back = FALSE,
  test_values = TRUE
)
```

```{r repeated-p-matrix}
rep_p_matrix <- result$outputs$repeated_p
colnames(rep_p_matrix) <- c("Analysis 1", "Analysis 2")
knitr::kable(rep_p_matrix, digits = 4,
             caption = "Repeated p-values at each analysis")
```

We now walk through the procedure analysis by analysis using repeated p-values.
At each analysis, the repeated p-values are passed to `graph_test_shortcut()`
for multiplicity adjustment. The shortcut procedure rejects hypotheses in
order of increasing adjusted repeated p-values. Newly rejected hypotheses are removed
from the graph before proceeding to the next analysis.

**Analysis 1.** The initial graph has weights $(0.5, 0.5, 0, 0)$. The repeated
p-values are all large (the OBF spending function allocates very little alpha
to the first interim), so no hypothesis is rejected. The graph is unchanged.

**Analysis 2.** The repeated p-values at this analysis are much smaller.
The graphical shortcut procedure rejects $H_1$ first, updates the graph to
propagate $H_1$'s weight, then rejects $H_2$, followed by $H_3$. $H_4$'s
repeated p-value remains too large, so it is retained. The rejection sequence
at analysis 2 is: $H_1 \to H_2 \to H_3$.

The printed result summarizes these steps. The line
`Group sequential engine: gsDesign` records that the group sequential
quantities were obtained from gsDesign, and the spending functions are
labeled with gsDesign's own names:

```{r print-gsd-no-lb}
print(result)
```

A hypothesis is rejected when its adjusted repeated p-value is at most
$\alpha = 0.025$. The adjusted repeated p-values at analysis 2 are
`r round(result$outputs$adjusted_p[["H1"]], 4)` ($H_1$),
`r round(result$outputs$adjusted_p[["H2"]], 4)` ($H_2$),
`r round(result$outputs$adjusted_p[["H3"]], 4)` ($H_3$), and
`r round(result$outputs$adjusted_p[["H4"]], 4)` ($H_4$).
The rejection sequence follows increasing adjusted repeated p-values:
$H_1$ (smallest) $\to H_2 \to H_3$. $H_4$'s adjusted repeated p-value
exceeds $\alpha$, so it is not rejected. All three rejections occur at
analysis 2.

#### Nominal boundaries

The same rejection decisions can be understood by comparing observed p-values
against nominal boundaries at each analysis. The `test_values` output from
`graph_test_shortcut_gsDesign()` provides these boundaries. Each boundary is
the upper bound of a gsDesign design constructed at the level currently
allocated to the hypothesis (its weight times $\alpha$), converted to the
p-value scale. The rejection sequence is determined by increasing adjusted
repeated p-values. After each rejection, the graph is updated and boundaries
for remaining hypotheses are recomputed with their new (increased) weights,
potentially enabling further rejections at the same analysis.

```{r test-values-tables}
format_test_values <- function(tv) {
  tv$Boundary <- formatC(tv$Boundary, format = "f", digits = 6)
  tv
}
knitr::kable(format_test_values(result$test_values[[1]]), digits = 6,
             caption = "Analysis 1: nominal boundaries and rejection decisions")
knitr::kable(format_test_values(result$test_values[[2]]), digits = 6,
             caption = "Analysis 2: nominal boundaries and rejection decisions")
```

**Analysis 1 (t = 1/3).** The initial weights are $(0.5, 0.5, 0, 0)$. The
OBF spending function allocates very little alpha to the first interim
analysis — as shown in the Analysis 1 table above, the nominal boundary for
$H_1$ and $H_2$ is approximately
`r formatC(result$test_values[[1]]$Boundary[1], format = "f", digits = 6)`.
Since both observed p-values (0.0062 for $H_1$ and 0.017 for $H_2$) exceed
this boundary, no hypothesis is rejected.

An important note: the nominal boundary computed at a fraction of alpha is
**not** equal to the same fraction of the boundary computed at the full alpha.
For example, the boundary at the first analysis with the OBF spending function:

```{r key-inequality}
design_half <- gsDesign(
  k = 3, test.type = 1, alpha = 0.0125,
  sfu = sfLDOF, n.I = c(1/3, 2/3, 1), maxn.IPlan = 1
)
cat(sprintf(
  "Boundary at alpha = 0.0125:     %.6f\n0.5 * Boundary at alpha = 0.025: %.6f\n",
  pnorm(design_half$upper$bound[1], lower.tail = FALSE),
  0.5 * pnorm(design_obf$upper$bound[1], lower.tail = FALSE)
))
```

This demonstrates why the design must be constructed at the
hypothesis-specific significance level (weight times alpha), rather than
applying the weight to boundaries computed at the full alpha. This is why
`graph_test_shortcut_gsDesign()` rebuilds the gsDesign design of a hypothesis
whenever its weight changes.

**Analysis 2 (t = 2/3).** The test_values table above shows the boundary for
each hypothesis at the point when it is tested, reflecting sequential graph
updates within the analysis:

- $H_1$ (weight 0.5): boundary =
  `r sprintf("%.4f", result$test_values[[2]]$Boundary[1])`.
  Since $p_{1,2} = 0.0002$ is below this boundary, $H_1$ is rejected. The
  graph is updated, propagating $H_1$'s weight.
- $H_2$ (weight updated to 0.75 after $H_1$ rejection): boundary =
  `r sprintf("%.4f", result$test_values[[2]]$Boundary[2])`.
  Since $p_{2,2} = 0.0035$ is below this boundary, $H_2$ is rejected. The
  graph is updated.
- $H_3$ (weight updated to 0.5 after $H_1$ and $H_2$ rejection): boundary =
  `r sprintf("%.4f", result$test_values[[2]]$Boundary[3])`.
  Since $p_{3,2} = 0.002$ is below this boundary, $H_3$ is rejected. The
  graph is updated.
- $H_4$ (weight updated to 1.0 after all three rejections): boundary =
  `r sprintf("%.4f", result$test_values[[2]]$Boundary[4])`.
  Since $p_{4,2} = 0.06$ exceeds this boundary, $H_4$ is retained.

Three out of four hypotheses ($H_1$, $H_2$, $H_3$) are rejected at
analysis 2.

#### Comparison with Paper Values

We can verify our results against Table 2 of Maurer and Bretz (2013). The
paper reports repeated p-values (which they call $\tilde{p}^r$) at each
analysis:

```{r paper-comparison, include=FALSE}
paper_repeated_p <- data.frame(
  Hypothesis = c("H1", "H2", "H3", "H4"),
  `pr_1 (paper)` = c(0.1141, 0.1683, 0.1316, 0.382),
  `pr_1 (ours)` = round(result$outputs$sequential_p[, 1], 4),
  `pr_2 (paper)` = c(0.0024, 0.0172, 0.0117, 0.1285),
  `pr_2 (ours)` = round(result$outputs$sequential_p[, 2], 4),
  check.names = FALSE
)
knitr::kable(paper_repeated_p, digits = 4,
             caption = "Comparison of sequential p-values with paper Table 2")
```

The nominal significance levels at each analysis also match the paper's Table 1:

```{r nominal-comparison, include=FALSE}
paper_table1 <- data.frame(
  Hypothesis = c("H1", "H2", "H3", "H4"),
  `alpha*_1 (paper)` = c(0.00002, 0.00002, 0, 0),
  `alpha*_1 (ours)` = round(result$test_values[[1]]$Boundary, 5),
  `alpha*_2 (paper)` = c(0.0022, "0.004 (updated)", "0.0022 (updated)",
                         "0.006 (updated)"),
  `alpha*_2 (ours)` = round(result$test_values[[2]]$Boundary, 4),
  check.names = FALSE
)
knitr::kable(paper_table1,
             caption = "Comparison of nominal boundaries with paper Table 1")
```

#### Boundary table

Setting `verbose = TRUE` produces a boundary table that lists the nominal
p-value boundaries at each analysis for all possible hypothesis weights from
the graph's closure. This table is independent of the observed p-values and
can be used to manually verify the rejection decisions shown in `test_values`.

```{r boundary-table}
result_verbose <- graph_test_shortcut_gsDesign(
  graph = g,
  p = p,
  alpha = 0.025,
  info_frac = info_frac,
  sfu = sfLDOF,
  verbose = TRUE
)

# Display boundary tables for each hypothesis
for (h in names(result_verbose$boundary_table)) {
  cat(h, "\n")
  print(result_verbose$boundary_table[[h]])
  cat("\n")
}
```

To verify the rejection decisions at analysis 2, look up each hypothesis's
weight in the boundary table and compare the `Boundary.2` column against
the observed p-value. For example, $H_1$ is tested at weight 0.5 with
boundary `r formatC(result_verbose$boundary_table$H1$Boundary.2[result_verbose$boundary_table$H1$Weight == 0.5], format = "f", digits = 7)`.
Since $p_{1,2} = 0.0002$ is below this boundary, $H_1$ is rejected. After
$H_1$'s rejection, the graph propagates weight to $H_2$ (weight 0.75) and
$H_3$ (weight 0.25). Looking up these weights confirms the boundaries used
in the `test_values` output.

## Case Study using sequential p-values and look_back = TRUE

This example is adapted from the
[gMCPLite vignette](https://cran.r-project.org/package=gMCPLite/vignettes/GraphicalMultiplicity.html)
for an oncology trial with three endpoints, two populations and two interim
analyses. Different hypotheses are tested at different analyses.

### Trial Setup

An oncology trial tests three endpoints — overall survival (OS), progression-free
survival (PFS), and objective response rate (ORR) — each in a biomarker
subgroup (S) and in the overall population (A, for all subjects). This gives
six hypotheses with different numbers of analyses:

- $H_{1}$ (OS, S): OS in the subgroup (3 analyses)
- $H_{2}$ (OS, A): OS in all subjects (3 analyses)
- $H_{3}$ (PFS, S): PFS in the subgroup (2 analyses)
- $H_{4}$ (PFS, A): PFS in all subjects (2 analyses)
- $H_{5}$ (ORR, S): ORR in the subgroup (1 analysis)
- $H_{6}$ (ORR, A): ORR in all subjects (1 analysis)

OS endpoints require longer follow-up and have three planned analyses. PFS
endpoints mature faster and have two analyses. ORR is assessed at a single
analysis. Each hypothesis has its own event-driven information fractions.

### Graph and Alpha Allocation

The initial alpha is split as: $H_1$ and $H_2$ each receive 0.01, $H_3$
receives 0.004, $H_4$ receives 0, $H_5$ and $H_6$ each receive 0.0005. This
totals $\alpha = 0.025$. The hypothesis weights are obtained by dividing each
allocation by $\alpha$:

```{r oncology-graph}
alpha_onc <- 0.025
alpha_allocation <- c(0.01, 0.01, 0.004, 0, 0.0005, 0.0005)
hypotheses_onc <- alpha_allocation / alpha_onc

hyp_names_onc <- c(
  "H1_OS_S", "H2_OS_A", "H3_PFS_S", "H4_PFS_A", "H5_ORR_S", "H6_ORR_A"
)
names(hypotheses_onc) <- hyp_names_onc

transitions_onc <- rbind(
  c(0, 1, 0, 0, 0, 0),       # H1_OS_S  -> H2_OS_A
  c(0, 0, 0.5, 0.5, 0, 0),   # H2_OS_A  -> H3_PFS_S (1/2), H4_PFS_A (1/2)
  c(0, 0, 0, 1, 0, 0),       # H3_PFS_S -> H4_PFS_A
  c(0, 0, 0, 0, 0.5, 0.5),   # H4_PFS_A -> H5_ORR_S (1/2), H6_ORR_A (1/2)
  c(0, 0, 0, 0, 0, 1),       # H5_ORR_S -> H6_ORR_A
  c(0.5, 0.5, 0, 0, 0, 0)    # H6_ORR_A -> H1_OS_S (1/2), H2_OS_A (1/2)
)

g_onc <- graph_create(hypotheses_onc, transitions_onc)
```

The transition structure follows the hierarchy: within each population, alpha
flows from OS to PFS to ORR, and ORR recycles to OS. Between populations,
the all-subjects hypotheses share alpha with the subgroup hypotheses.

```{r oncology-graph-plot, eval = requireNamespace("igraph", quietly = TRUE), fig.height=5.5, fig.width=6, out.width="100%"}
onc_layout <- rbind(
  c(0, 3),     # H1_OS_S
  c(2, 3),     # H2_OS_A
  c(0, 1.8),   # H3_PFS_S
  c(3, 1.8),   # H4_PFS_A
  c(0, 0.5),   # H5_ORR_S
  c(2, 0.5)    # H6_ORR_A
)
# Edge label positions: NA = auto, explicit coords to move specific labels
# Edge order: 1=H6->H1, 2=H1->H2, 3=H6->H2, 4=H2->H3, 5=H2->H4,
#             6=H3->H4, 7=H4->H5, 8=H4->H6, 9=H5->H6
label_x <- rep(NA, 9)
label_y <- rep(NA, 9)
label_x[1] <- 0.4;  label_y[1] <- 2.5    # H6->H1: toward arrow (H1)
label_x[3] <- 2.0;  label_y[3] <- 2.375  # H6->H2: on the edge, toward arrow
label_x[4] <- 1.5;  label_y[4] <- 2.7    # H2->H3: toward tail (H2)
label_x[6] <- 0.75; label_y[6] <- 1.8    # H3->H4: toward tail (H3)
label_x[7] <- 0.9;  label_y[7] <- 0.89   # H4->H5: toward arrow (H5)
label_x[8] <- 2.5;  label_y[8] <- 1.15   # H4->H6: on the edge, midway

plot(g_onc, layout = onc_layout, vertex.size = 60, asp = 1,
     vertex.label.cex = 0.7,
     rescale = FALSE,
     xlim = c(-0.6, 3.6),
     ylim = c(-0.1, 3.6),
     edge.label.x = label_x,
     edge.label.y = label_y)
```

### P-values and Information Fractions

Since hypotheses have different numbers of analyses, we use `NA` padding.
The p-value and information fraction matrices have $K = 3$ columns (the
maximum number of analyses), with `NA` for analyses that do not apply.

When `p` contains `NA` values, `info_frac` must be a matrix with `NA` in the
same positions. This ensures users explicitly specify the correct information
fractions for each hypothesis, since different endpoints may have very
different event accrual schedules. Internally, `graph_test_shortcut_gsDesign()`
builds one gsDesign design per hypothesis from its non-`NA` analyses, so a
hypothesis with three analyses gets a three-analysis design, a hypothesis with
two analyses a two-analysis design, and a hypothesis with a single analysis
is handled as a fixed-sample test (gsDesign requires at least two analyses).

```{r oncology-p-values}
p_onc <- rbind(
  H1_OS_S  = c(0.03,    0.0001, 0.000001),
  H2_OS_A  = c(0.2,     0.15,   0.1),
  H3_PFS_S = c(0.2,     0.001,  NA),
  H4_PFS_A = c(0.3,     0.2,    NA),
  H5_ORR_S = c(0.00001, NA,     NA),
  H6_ORR_A = c(0.1,     NA,     NA)
)

# Event-driven information fractions (events / max events per endpoint)
info_frac_onc <- rbind(
  H1_OS_S  = c(185 / 295, 245 / 295, 1),
  H2_OS_A  = c(529 / 800, 700 / 800, 1),
  H3_PFS_S = c(265 / 310, 1,         NA),
  H4_PFS_A = c(675 / 750, 1,         NA),
  H5_ORR_S = c(1,         NA,        NA),
  H6_ORR_A = c(1,         NA,        NA)
)

knitr::kable(
  data.frame(
    Hypothesis = rownames(p_onc),
    `No. of Analyses` = rowSums(!is.na(p_onc)),
    `Info Frac` = apply(info_frac_onc, 1, function(x) {
      paste(round(x[!is.na(x)], 3), collapse = ", ")
    }),
    check.names = FALSE
  ),
  row.names = FALSE,
  caption = "Number of analyses and information fractions per hypothesis"
)
```

```{r oncology-p-table}
p_display <- p_onc
p_display[] <- ifelse(
  is.na(p_display), "—",
  formatC(as.numeric(p_display), format = "f", digits = 6)
)
knitr::kable(
  as.data.frame(p_display),
  col.names = paste("Analysis", 1:3),
  caption = "Observed nominal p-values (— indicates no data at that analysis)"
)
```

### Sequential P-values (look_back = TRUE)

The gMCPLite vignette uses sequential p-values for its analysis. We replicate
this by running `graph_test_shortcut_gsDesign()` with `look_back = TRUE`,
which uses sequential p-values for rejection decisions. All hypotheses use
the Lan-DeMets O'Brien-Fleming spending function:

```{r oncology-run-lb}
result_onc_lb <- graph_test_shortcut_gsDesign(
  graph = g_onc,
  p = p_onc,
  alpha = alpha_onc,
  info_frac = info_frac_onc,
  sfu = sfLDOF,
  look_back = TRUE,
  test_values = TRUE
)
print(result_onc_lb)
```

The sequential p-value matrices show the full picture across analyses, with
`NA` for hypothesis–analysis combinations that do not exist:

```{r oncology-seq-p}
seq_p_display <- result_onc_lb$outputs$sequential_p
seq_p_display[] <- ifelse(
  is.na(seq_p_display), NA,
  formatC(as.numeric(seq_p_display), format = "f", digits = 6)
)
knitr::kable(
  as.data.frame(seq_p_display),
  caption = "Sequential p-values (NA = hypothesis not tested at that analysis)"
)
```

The procedure processes three analyses sequentially:

**Analysis 1.** Hypotheses $H_1$–$H_4$ are tested at their first interim
analysis. $H_5$ and $H_6$ undergo their only analysis. The ORR subgroup
hypothesis $H_5$ has a very small p-value ($p = 0.00001$) and is rejected.
Its alpha is propagated to $H_6$ through the graph.

**Analysis 2.** Hypotheses $H_1$ and $H_2$ are at their second interim.
$H_3$ and $H_4$ are at their final analysis. $H_5$ and $H_6$ have no further
analyses (their columns are `NA`). $H_1$ (OS subgroup) and $H_3$ (PFS
subgroup) cross their boundaries and are rejected.

**Analysis 3.** Only $H_1$ and $H_2$ have data at the final analysis, but
$H_1$ was already rejected. $H_2$'s p-values remain too large throughout.

```{r oncology-results-lb}
# Extract the p-value and boundary at the decision point for each hypothesis.
# For rejected hypotheses, this is the analysis where rejection occurred.
# For non-rejected hypotheses, this is the last analysis where it was tested.
tv_all <- do.call(rbind, result_onc_lb$test_values)
decision_p <- numeric(length(hyp_names_onc))
decision_bound <- numeric(length(hyp_names_onc))
for (i in seq_along(hyp_names_onc)) {
  h <- hyp_names_onc[i]
  rows <- tv_all[tv_all$Hypothesis == h, ]
  if (nrow(rows) == 0) next
  # Use the last row for this hypothesis (final decision point)
  decision_p[i] <- rows$p[nrow(rows)]
  decision_bound[i] <- rows$Boundary[nrow(rows)]
}

# For non-rejected hypotheses, find the last analysis where they were tested
last_tested <- rowSums(!is.na(p_onc))

onc_summary_lb <- data.frame(
  Hypothesis = hyp_names_onc,
  `p` = formatC(decision_p, format = "f", digits = 6),
  Boundary = formatC(decision_bound, format = "f", digits = 6),
  Rejected = result_onc_lb$outputs$rejected,
  Tested.at = result_onc_lb$outputs$decision_at,
  First.Rej.at = ifelse(
    is.na(result_onc_lb$outputs$first_rejected_at), "—",
    as.character(result_onc_lb$outputs$first_rejected_at)
  ),
  Last.Rej.at = ifelse(
    is.na(result_onc_lb$outputs$last_rejected_at), "—",
    as.character(result_onc_lb$outputs$last_rejected_at)
  ),
  check.names = FALSE
)
knitr::kable(onc_summary_lb, row.names = FALSE,
             caption = "Oncology case study (look_back = TRUE): rejection decisions")
```

This case study demonstrates that `graph_test_shortcut_gsDesign()` handles
trials where different endpoints have different numbers of analyses — a
common situation in oncology trials with OS, PFS, and ORR endpoints.

### Per-hypothesis look_back

The `look_back` parameter can also be specified as a logical vector, allowing
different hypotheses to use different testing strategies. This is useful when
some endpoints benefit from looking back at earlier evidence while others
do not.

Consider a scenario where we want to use `look_back = TRUE` for ORR
endpoints (to allow them to benefit from earlier evidence if they receive
alpha via graph propagation at a later analysis) while using
`look_back = FALSE` for OS and PFS endpoints:

```{r per-hyp-look-back}
look_back_onc <- c(
  FALSE,  # H1_OS_S:  no look_back
  FALSE,  # H2_OS_A:  no look_back
  FALSE,  # H3_PFS_S: no look_back
  FALSE,  # H4_PFS_A: no look_back
  TRUE,   # H5_ORR_S: look_back
  TRUE    # H6_ORR_A: look_back
)

result_onc_mixed <- graph_test_shortcut_gsDesign(
  graph = g_onc,
  p = p_onc,
  alpha = alpha_onc,
  info_frac = info_frac_onc,
  sfu = sfLDOF,
  look_back = look_back_onc
)
print(result_onc_mixed)
```

```{r per-hyp-look-back-compare}
mixed_comparison <- data.frame(
  Hypothesis = hyp_names_onc,
  `Rejected (all LB)` = result_onc_lb$outputs$rejected,
  `Adj. P (all LB)` = round(result_onc_lb$outputs$adjusted_p, 6),
  `Rejected (mixed)` = result_onc_mixed$outputs$rejected,
  `Adj. P (mixed)` = round(result_onc_mixed$outputs$adjusted_p, 6),
  check.names = FALSE
)
knitr::kable(mixed_comparison, row.names = FALSE,
             caption = "Effect of per-hypothesis look_back on rejection decisions")
```

In this example, using `look_back = FALSE` for OS and PFS means their
rejection decisions are based solely on repeated p-values at each analysis.
The ORR hypotheses use `look_back = TRUE`, allowing them to benefit from
evidence at earlier analyses if they receive alpha via graph propagation
at a later analysis. This mixed strategy can be useful when different
endpoints have different clinical or regulatory rationales for looking back.

### When look_back makes a difference

We modify the oncology trial p-values to illustrate a scenario where
`look_back = TRUE` leads to additional rejections. Three key changes are
made:

- $H_2$ (OS, all subjects): strong evidence at both analysis 1
  ($p = 0.003$) and analysis 2 ($p = 0.005$). At analysis 1, $H_2$'s
  weight (0.4) gives a boundary of 0.00154 — too small for $p = 0.003$.
  At analysis 2, after $H_1$ is rejected, $H_2$ receives $H_1$'s weight
  (reaching 0.8) and its analysis-2 p-value crosses the boundary.
  Additionally, looking back to analysis 1 with the increased weight gives
  a boundary of 0.00423, which $p = 0.003$ also crosses — so $H_2$ is
  attributed to analysis 1.
- $H_4$ (PFS, all subjects): strong evidence at analysis 1
  ($p = 0.0001$) but weaker at analysis 2 ($p = 0.02$). Since $H_4$ starts
  with weight 0, it cannot be tested until it receives alpha via graph
  propagation — which only happens after $H_3$ is rejected at analysis 2.
- $H_5$ (ORR, subgroup): a single analysis with $p = 0.0008$, which is
  above its initial boundary ($0.0005$) but below the boundary it would
  receive after graph propagation from $H_4$'s rejection.

```{r oncology-look-back-diff}
p_onc_lb <- p_onc
p_onc_lb["H2_OS_A", ] <- c(0.003, 0.005, 0.1)
p_onc_lb["H4_PFS_A", ] <- c(0.0001, 0.02, NA)
p_onc_lb["H5_ORR_S", ] <- c(0.0008, NA, NA)

# look_back = FALSE: H4's repeated p-value at analysis 2 is too large
result_onc_no_lb <- graph_test_shortcut_gsDesign(
  graph = g_onc,
  p = p_onc_lb,
  alpha = alpha_onc,
  info_frac = info_frac_onc,
  sfu = sfLDOF,
  look_back = FALSE,
  test_values = TRUE
)

# look_back = TRUE: H4 benefits from its strong analysis-1 evidence
result_onc_yes_lb <- graph_test_shortcut_gsDesign(
  graph = g_onc,
  p = p_onc_lb,
  alpha = alpha_onc,
  info_frac = info_frac_onc,
  sfu = sfLDOF,
  look_back = TRUE,
  test_values = TRUE
)
```

With `look_back = FALSE`, $H_4$ and $H_5$ are not rejected, while $H_2$ is
rejected at analysis 2:

```{r oncology-look-back-diff-no}
print(result_onc_no_lb)
```

With `look_back = TRUE`, $H_2$, $H_4$, and $H_5$ **are all rejected** and
attributed to analysis 1. The `test_values` output at analysis 2 shows
three different look_back scenarios:

- $H_2$: its nominal p-value at analysis 2 ($p = 0.005$) crosses the
  boundary at analysis 2 after receiving $H_1$'s weight — so $H_2$ is
  rejected at analysis 2 regardless of look_back. But looking back to
  analysis 1 with the increased weight (0.8) gives a boundary of 0.00423,
  which $p = 0.003$ also crosses. This means $H_2$ is rejected at both
  analyses, and is attributed to analysis 1 (`First.Rej.at = 1`). In the
  `test_values`, both the analysis-2 row and the look_back analysis-1 row
  show `Reject = TRUE`.
- $H_4$: its nominal p-value at analysis 2 ($p = 0.02$) does not cross the
  boundary, but looking back to analysis 1, $p = 0.0001$ crosses the
  boundary with the weight received after $H_3$'s rejection.
- $H_5$: it has no data at analysis 2 (single-analysis endpoint), but
  looking back to analysis 1, $p = 0.0008$ crosses the boundary with the
  weight received after $H_4$'s rejection. Note that $H_5$ only appears
  as a look_back row (marked with `*`) in the analysis 2 test_values,
  since it has no analysis-2 data.

```{r oncology-look-back-diff-yes}
print(result_onc_yes_lb)
```

```{r oncology-look-back-diff-compare}
lb_diff_table <- data.frame(
  Hypothesis = hyp_names_onc,
  `Rejected (no LB)` = result_onc_no_lb$outputs$rejected,
  `Rejected (LB)` = result_onc_yes_lb$outputs$rejected,
  `First.Rej.at (LB)` = ifelse(
    is.na(result_onc_yes_lb$outputs$first_rejected_at), "--",
    as.character(result_onc_yes_lb$outputs$first_rejected_at)),
  `Last.Rej.at (LB)` = ifelse(
    is.na(result_onc_yes_lb$outputs$last_rejected_at), "--",
    as.character(result_onc_yes_lb$outputs$last_rejected_at)),
  check.names = FALSE
)
knitr::kable(lb_diff_table, row.names = FALSE,
             caption = "Effect of look_back on oncology trial rejection decisions")
```

The three look_back rejections illustrate different mechanisms:

- $H_2$ had data at both analyses and was rejected at analysis 2 even
  without look_back. But graph propagation from $H_1$'s rejection increased
  $H_2$'s weight, and looking back to analysis 1 with the new weight showed
  that the boundary was also crossed there — attributing the rejection to
  the earlier analysis.
- $H_4$ started with weight 0 and only became testable after $H_3$'s
  rejection at analysis 2. With `look_back = FALSE`, $H_4$ could only use
  its analysis-2 evidence ($p = 0.02$), which was insufficient. With
  `look_back = TRUE`, the strong analysis-1 evidence ($p = 0.0001$)
  crossed the boundary.
- $H_5$ had a single analysis and was not rejected initially (boundary too
  small). After $H_4$'s rejection propagated weight to $H_5$, looking back
  to analysis 1 with the increased weight enabled rejection — even though
  $H_5$ had no data at analysis 2.

### Customizing Spending Functions: Spending Time

gsDesign separates *spending time* from *information fraction*. The
information fraction (`n.I` in gsDesign, `info_frac` here) determines the
correlation structure of the test statistics, while the spending time
(`usTime`) determines how alpha is allocated across analyses via the
spending function. The two can differ when, for example, all-subjects
hypotheses use all-subjects event counts for the correlation but subgroup
event counts for spending.

`graph_test_shortcut_gsDesign()` passes gsDesign's `usTime` argument through
unchanged. By default (`usTime = NULL`), the spending time equals the
information fraction (capped at 1). Otherwise, `usTime` is either a vector
with one value per analysis, applied to all hypotheses, or a matrix with the
same dimensions and `NA` positions as `info_frac`, giving each hypothesis its
own spending time. This plays the role of `spending_with_time()` in
`graph_test_shortcut_gsd()`.

Consider the oncology trial above. The all-subjects hypotheses ($H_2$ and
$H_4$) use all-subjects event counts for their information fractions (which
determine the correlation structure), but one might want to use the
corresponding subgroup event counts as the spending time (which determines
how aggressively alpha is spent at each analysis). This is because the
subgroup is a subset of the all-subjects population, and the subgroup events
may better reflect the information available for the treatment effect
comparison.

For the oncology trial, $H_2$ (OS, all subjects) uses all-subjects OS events
(`r paste(c(529, 700, 800), collapse = ", ")`) for the correlation structure
(via `info_frac`), but subgroup OS events
(`r paste(c(185, 245, 295), collapse = ", ")`) for spending. We start from
the information fractions, so that the `NA` positions match, and replace the
rows of the all-subjects hypotheses:

```{r spending-time-setup}
usTime_onc <- info_frac_onc

# H2 (OS, all subjects): spending time = subgroup OS event fractions
usTime_onc["H2_OS_A", ] <- c(185 / 295, 245 / 295, 1)

# H4 (PFS, all subjects): spending time = subgroup PFS event fractions
usTime_onc["H4_PFS_A", ] <- c(265 / 310, 1, NA)

usTime_onc
```

Now `info_frac_onc` continues to use all-subjects event counts for $H_2$ and
$H_4$ (determining the correlation structure), while `usTime_onc` uses
subgroup event counts for alpha allocation. The printed result shows the
spending time alongside the information fractions:

```{r spending-time-run}
result_onc_st <- graph_test_shortcut_gsDesign(
  graph = g_onc,
  p = p_onc,
  alpha = alpha_onc,
  info_frac = info_frac_onc,
  sfu = sfLDOF,
  usTime = usTime_onc,
  look_back = TRUE
)
print(result_onc_st)
```

```{r spending-time-compare}
st_comparison <- data.frame(
  Hypothesis = hyp_names_onc,
  `Rejected (info fraction)` = result_onc_lb$outputs$rejected,
  `Adj. P (info fraction)` = round(result_onc_lb$outputs$adjusted_p, 6),
  `Rejected (spending time)` = result_onc_st$outputs$rejected,
  `Adj. P (spending time)` = round(result_onc_st$outputs$adjusted_p, 6),
  check.names = FALSE
)
knitr::kable(st_comparison, row.names = FALSE,
             caption = "Effect of spending time on rejection decisions")
```

The spending time adjustment affects the sequential p-values for $H_2$ and
$H_4$ because it changes how alpha is allocated across their analyses. The
subgroup event fractions are smaller than the all-subjects event fractions at
early analyses (e.g., `r round(185/295, 3)` vs. `r round(529/800, 3)` at
analysis 1 for OS), meaning the spending function allocates less alpha to
early analyses — the boundaries become more conservative at interim analyses
and more liberal at the final analysis.

Internally, the spending time is passed to both `gsDesign()` (for the
boundaries) and `sequentialPValue()` (for the repeated p-values), so the two
views of the procedure remain consistent.

### Monitoring: Adjusting for Changed Final Information

In practice, the total information (e.g., total number of events) at the
final analysis may differ from what was planned at the design stage. When
this happens, the information fractions at earlier analyses change
retroactively — not because the data changed, but because the denominator
(planned total) is now different. This creates a challenge: the boundaries
at earlier analyses were already computed using the **planned** information
fractions, but the correlation structure at the current analysis should
reflect the **actual** information fractions.

The separation of spending time and information fraction handles this
naturally:

- **Spending time** (`usTime`): use the planned information fractions to
  preserve the boundaries at analyses 1 and 2 (which have already been
  applied).
- **Information fraction** (`info_frac`): use the actual information fractions
  for the correlation structure, since this reflects the true joint
  distribution of the test statistics.

Consider the OS subgroup hypothesis ($H_1$) in the oncology trial. The
trial was designed with a planned maximum of 295 OS events in the subgroup,
giving planned information fractions of
(`r paste(round(c(185, 245, 295) / 295, 3), collapse = ", ")`).
Suppose that by the time of the final analysis, 310 events have been
observed instead of 295. The actual information fractions are now
(`r paste(round(c(185, 245, 310) / 310, 3), collapse = ", ")`):

```{r monitoring-setup}
# Planned info fractions (used for boundaries at analyses 1 and 2)
planned_if_h1 <- c(185 / 295, 245 / 295, 1)

# Actual info fractions (more events than planned at final analysis)
actual_if_h1 <- c(185 / 310, 245 / 310, 1)

cat("Planned:", round(planned_if_h1, 4), "\n")
cat("Actual: ", round(actual_if_h1, 4), "\n")
```

To illustrate, we compare the boundaries computed by `gsDesign()` under three
scenarios:

1. **Planned**: both spending and correlation use planned info fractions
   (the original design).
2. **Naive update**: both spending and correlation use actual info fractions
   (ignores that analyses 1 and 2 used planned boundaries).
3. **Correct monitoring**: spending uses planned info fractions (`usTime`),
   correlation uses actual info fractions (`n.I`).

```{r monitoring-comparison}
alpha_h1 <- 0.01  # H1's allocated alpha

# Nominal p-value boundaries of a gsDesign design with information n_i and
# spending time us_time
nominal_bounds <- function(n_i, us_time) {
  design <- gsDesign(
    k = 3, test.type = 1, alpha = alpha_h1,
    sfu = sfLDOF, n.I = n_i, maxn.IPlan = 1, usTime = us_time
  )
  pnorm(design$upper$bound, lower.tail = FALSE)
}

# Scenario 1: Planned (original design)
bounds_planned <- nominal_bounds(planned_if_h1, planned_if_h1)

# Scenario 2: Naive update (both use actual)
bounds_naive <- nominal_bounds(actual_if_h1, actual_if_h1)

# Scenario 3: Correct monitoring (spending=planned, correlation=actual)
bounds_monitor <- nominal_bounds(actual_if_h1, planned_if_h1)

monitor_table <- data.frame(
  Analysis = 1:3,
  Planned.IF = round(planned_if_h1, 4),
  Actual.IF = round(actual_if_h1, 4),
  Boundary.Planned = bounds_planned,
  Boundary.Naive = bounds_naive,
  Boundary.Monitor = bounds_monitor
)
knitr::kable(monitor_table, digits = 6,
             caption = paste("Boundaries under three scenarios",
                             "(H1 with alpha = 0.01)"))
```

The key observations:

- **Analyses 1 and 2**: the monitoring boundaries coincide with the planned
  boundaries. The spending at these analyses is preserved (the same
  cumulative alpha is allocated), and since the information at analyses 1 and
  2 has not changed relative to each other, the correlation between them is
  unchanged as well.
- **Analysis 3**: the monitoring boundary reflects both the preserved
  spending schedule and the updated correlation structure.
- **Naive update**: changes the spending at all analyses, which is
  inconsistent with the boundaries already applied at analyses 1 and 2.

This approach extends naturally to the full graphical procedure. For
monitoring at analysis 3, use the actual information fractions in
`info_frac` and the planned information fractions in `usTime`:

```{r monitoring-full}
# Actual info fractions for all hypotheses
# (only OS hypotheses affected; PFS and ORR are complete)
actual_if_onc <- info_frac_onc
actual_if_onc["H1_OS_S", ] <- c(185 / 310, 245 / 310, 1)
actual_if_onc["H2_OS_A", ] <- c(529 / 830, 700 / 830, 1)

# Planned info fractions as the spending time for all hypotheses
result_monitor <- graph_test_shortcut_gsDesign(
  graph = g_onc,
  p = p_onc,
  alpha = alpha_onc,
  info_frac = actual_if_onc,
  sfu = sfLDOF,
  usTime = info_frac_onc,
  look_back = TRUE
)
print(result_monitor)
```

An equivalent way to describe the same situation, common in gsDesign, keeps
the planned maximum as the unit of information, so that the information
fraction at the final analysis exceeds 1 (e.g., $310 / 295 = 1.051$ for
$H_1$). `graph_test_shortcut_gsDesign()` accepts information fractions above
1, and the default spending time — the information fractions capped at 1 — is
then exactly the planned schedule. Since the correlation between analyses
depends only on ratios of information, this gives the same results:

```{r monitoring-overrun}
overrun_if_onc <- info_frac_onc
overrun_if_onc["H1_OS_S", ] <- c(185, 245, 310) / 295
overrun_if_onc["H2_OS_A", ] <- c(529, 700, 830) / 800

result_overrun <- graph_test_shortcut_gsDesign(
  graph = g_onc,
  p = p_onc,
  alpha = alpha_onc,
  info_frac = overrun_if_onc,
  sfu = sfLDOF,
  look_back = TRUE
)
all.equal(result_overrun$outputs$adjusted_p, result_monitor$outputs$adjusted_p)
```

One caveat applies when calling `sequentialPValue()` directly with
over-running information: its default spending time is the information
divided by the *last* information supplied, which rescales the planned
spending at the interim analyses. `graph_test_shortcut_gsDesign()` avoids
this by always passing the spending time explicitly.

## Additional Examples

### Different Spending Functions per Hypothesis

The procedure allows different hypotheses to use different spending functions
by passing a list to `sfu`, with one element per hypothesis. For example, one
might use a more aggressive spending function for secondary hypotheses:

```{r different-spending}
result2 <- graph_test_shortcut_gsDesign(
  graph = g,
  p = p,
  alpha = 0.025,
  info_frac = c(1/3, 2/3),
  sfu = list(
    sfLDOF,     # H1: O'Brien-Fleming
    sfLDOF,     # H2: O'Brien-Fleming
    sfLDPocock, # H3: Pocock (more aggressive)
    sfLDPocock  # H4: Pocock (more aggressive)
  ),
  test_values = TRUE
)
print(result2)
```

Likewise, `sfupar` can be a list with one parameter per hypothesis.

### gsDesign's Spending Function Library

Any of gsDesign's spending functions can be passed as `sfu`, with its
parameter as `sfupar`, and so can the classical boundary families that
gsDesign specifies by name: `"OF"` (O'Brien-Fleming), `"Pocock"`, and `"WT"`
(Wang-Tsiatis, with `sfupar` giving the shape parameter $\Delta$). gsDesign
represents these families by their boundary shape rather than by a spending
function; `graph_test_shortcut_gsDesign()` obtains their repeated p-values by
inverting the boundary directly, and the resulting Wang-Tsiatis boundaries are
exact. The following compares several choices on the diabetes trial:

```{r gsdesign-library}
spending_choices <- list(
  "sfLDOF" = list(sfu = sfLDOF, sfupar = NULL),
  "sfHSD, gamma = -2" = list(sfu = sfHSD, sfupar = -2),
  "sfPower, rho = 2" = list(sfu = sfPower, sfupar = 2),
  "sfExponential, nu = 0.75" = list(sfu = sfExponential, sfupar = 0.75),
  "sfLinear, 20% spent at t = 0.5" = list(sfu = sfLinear, sfupar = c(0.5, 0.2)),
  "WT, Delta = 0.25" = list(sfu = "WT", sfupar = 0.25)
)

library_comparison <- do.call(rbind, lapply(names(spending_choices), function(nm) {
  choice <- spending_choices[[nm]]
  res <- graph_test_shortcut_gsDesign(
    graph = g,
    p = p,
    alpha = 0.025,
    info_frac = c(1/3, 2/3),
    sfu = choice$sfu,
    sfupar = choice$sfupar
  )
  data.frame(
    Spending = nm,
    t(round(res$outputs$adjusted_p, 4)),
    Rejected = paste(names(which(res$outputs$rejected)), collapse = ", "),
    check.names = FALSE
  )
}))
knitr::kable(library_comparison, row.names = FALSE,
             caption = "Adjusted p-values and rejections by spending function")
```

Two details of the interface are worth noting:

- A spending function whose parameter is itself a list, such as
  `sfTruncated()`, must have that list wrapped in a list with one copy per
  hypothesis, since a list given as `sfupar` is otherwise read as one
  parameter per hypothesis. For example, truncated Hwang-Shih-DeCani spending
  that spends nothing before 20% and everything by 80% information:

```{r gsdesign-truncated}
truncated_param <- list(trange = c(0.2, 0.8), sf = sfHSD, param = 1)
result_truncated <- graph_test_shortcut_gsDesign(
  graph = g,
  p = p,
  alpha = 0.025,
  info_frac = c(1/3, 2/3),
  sfu = sfTruncated,
  sfupar = rep(list(truncated_param), 4)
)
result_truncated$outputs$rejected
```

- `sfPoints()`, which takes the cumulative proportion of alpha spent at each
  analysis, is not supported: gsDesign evaluates the spending function on the
  analyses observed so far when computing repeated p-values, and `sfPoints()`
  requires the full schedule at every call. The same spending is available
  through `sfLinear()`, whose parameter lists the analysis times followed by
  the cumulative proportions spent at them.

### User-Defined Spending Functions

Users can define entirely custom spending functions following gsDesign's
convention. A spending function takes three arguments — `alpha`, `t`, and
`param` — and returns a `spendfn` object whose `spend` element is the
cumulative alpha spent at each `t`. The easiest way to construct such an
object is gsDesign's template `spendingFunction()`, which fills in the
required structure; the custom function then only needs to replace the
`spend` element and, for informative printing, the `name`. gsDesign's own
spending functions also store a reference to themselves in the `sf` element,
which `gsDesign()` and `sequentialPValue()` use to evaluate the spending
function later; `graph_test_shortcut_gsDesign()` sets this element for the
functions it is given, but following the convention keeps a custom function
usable directly with gsDesign as well.

As an example, consider a **piecewise spending function** that uses
O'Brien-Fleming spending for the first half of alpha (conservative at early
analyses) and Pocock spending for the second half (more aggressive). This
can arise naturally in the graphical procedure: a hypothesis that starts
with a small weight uses conservative OBF spending, and when it receives
additional weight via graph propagation, the incremental alpha is spent
more aggressively using Pocock. The threshold is the spending function's
parameter:

```{r user-defined-spending}
# Piecewise spending: OBF for the first `param` of alpha,
# Pocock for the remainder
sfPiecewise <- function(alpha, t, param) {
  x <- spendingFunction(alpha, t, param)
  x$name <- "Piecewise O'Brien-Fleming / Pocock"
  x$parname <- "threshold"
  x$sf <- sfPiecewise
  x$spend <- sfLDOF(min(alpha, param), t)$spend +
    if (alpha > param) sfLDPocock(alpha - param, t)$spend else 0
  x
}

# At alpha = 0.025: first 0.0125 is OBF, second 0.0125 is Pocock
t <- c(1/3, 2/3, 1)
comparison <- data.frame(
  `Info Fraction` = t,
  `OBF only` = sfLDOF(0.025, t)$spend,
  `Pocock only` = sfLDPocock(0.025, t)$spend,
  Piecewise = sfPiecewise(0.025, t, param = 0.0125)$spend,
  check.names = FALSE
)
knitr::kable(comparison, digits = 6,
             caption = "Piecewise spending: OBF for first half, Pocock for second half")
```

At `alpha = 0.0125` (e.g., a hypothesis with weight 0.5), the piecewise
function reduces to pure O'Brien-Fleming:

```{r user-defined-spending-half}
# At alpha = 0.0125: entirely OBF (below threshold)
all.equal(
  sfPiecewise(0.0125, t, param = 0.0125)$spend,
  sfLDOF(0.0125, t)$spend
)
```

The piecewise spending function can be used directly in
`graph_test_shortcut_gsDesign()`, with the threshold passed as `sfupar`:

```{r user-defined-spending-gsd}
result_piecewise <- graph_test_shortcut_gsDesign(
  graph = g,
  p = p,
  alpha = 0.025,
  info_frac = c(1/3, 2/3),
  sfu = sfPiecewise,
  sfupar = 0.0125
)
print(result_piecewise)
```

## Summary

The `graph_test_shortcut_gsDesign()` function performs multiple testing in
group sequential trials using graphical approaches, with the group sequential
calculations delegated to gsDesign. Key features include:

1. **gsDesign's group sequential machinery**: For each hypothesis, a
   one-sided gsDesign design encodes its analysis schedule. Boundaries are
   the design's upper bounds at the level currently allocated to the
   hypothesis, and repeated p-values are obtained from `sequentialPValue()`
   by setting the Z-statistics of earlier analyses to $-20$. The
   probabilities are evaluated by gsDesign's deterministic numerical
   integration, so results are reproducible without setting a seed.

2. **gsDesign's spending specification**: Spending functions (`sfu`), their
   parameters (`sfupar`), and the spending time (`usTime`) are specified
   exactly as in `gsDesign()`, giving access to gsDesign's spending function
   library, the classical boundary families, and user-defined spending
   functions built on `spendingFunction()`.

3. **The same graphical procedure**: Sequential rejection within analyses,
   repeated and sequential p-values, the `look_back` option, per-hypothesis
   spending functions and information fractions, and different numbers of
   analyses per hypothesis (via `NA` padding) all work as in
   `graph_test_shortcut_gsd()`, and the two functions return results of the
   same structure.

4. **Spending time and monitoring**: `usTime` separates the spending time
   from the information fraction, which also handles a final analysis whose
   information differs from the plan.

## References

Anderson, K. M. (2026). gsDesign: Group Sequential Design. R package.
[https://CRAN.R-project.org/package=gsDesign](https://CRAN.R-project.org/package=gsDesign)

Maurer, W., and Bretz, F. (2013). Multiple testing in group sequential trials
using graphical approaches. *Statistics in Biopharmaceutical Research*, 5(4),
311-320.

Zhao, Y., Liu, Q., Sun, L. Z., and Anderson, K. M. (2025). Adjusted
inference for multiple testing procedure in group-sequential designs.
*Biometrical Journal*, 67(1), e70020.
[doi:10.1002/bimj.70020](https://doi.org/10.1002/bimj.70020)
